PoPoolation Pipeline

### 2022-01-13 poolseq.Rmd
fixing typos

I mostly followed the PoPoolation and PoPoolation2 pipelines by Robert Kofler and Christian Schlötterer (https://sourceforge.net/p/popoolation2/wiki/Main/). The website has an easy to follow tutorial for using PoPoolation2 and how to use their scripts. This pipeline extends the use of the scripts for running on a Linux cluster. I highly recommend following the PoPoolation2 tutorial and manual and using these scripts if you have any modifications from default flags.
*Note: These scripts were written for use on Louisiana State University’s high performance cluster “SuperMike-II” running Red Hat Enterprise Linux 6 (more info found here: http://www.hpc.lsu.edu/resources/hpc/system.php?system=SuperMike-II)

The following scripts were used to analyze data from an experimental evolution study on Tigriopus californicus copepods (Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289). Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/). To download this dataset, go to: https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=run_browser , then type in the accession number for the identified files you want to download (e.g., the accession no. for the file Bodega population file is SRR10760095, identified under the "Run" heading). Then click on the Data access tab and there will be a downloadable link. Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER

***Strengths and weaknesses of Pool-seq***
Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information. Therefore, analyses rely on allele frequency estimates rather than hard “genotype calls”. PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome. The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution). For example, to identify small changes in allele frequencies, you would neet genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding. However, individual genotype information is lost, as well as information on dominance and
effect sizes. For example, individual genotypes cannot be linked to individual phenotypes. Additionally, estimations of linkage disequilibrium are limited to a single sequencing read since that is the known limit of haplotype information for a single individual. Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited. Unlike sequencing of individuals, this cannot be solved by analysing multiple reads from the same region of a single chromosome. One way to ameliorate this issue is to sequencing replicates of pools. In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods. Care must be taken to ensure the the experimentla design maximizes the ratio sequencing coverage to the number of individuals pooled.

***Notes on Experimental Design for Downstream Analyses:***
It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample. Robert Kofler and team also recommend that the pool size should be larger than the goal coverage for sequencing, because this minimizes re-sampling the same allele from a single individual several times. Because PoPoolation is designed for comparing differences in allele frequencies among treatments/population, accuracy will depend on the pool size and sequencing coverage. For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage. The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a singificant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient. If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
The sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig. 1a from Schlötterer et al., 2014).

@@

Customizeable Filtering and Assessing Quality of Pool-Seq Data:
https://github.com/ToBoDev/assessPool
This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data. It determines pool number and prepares proper data structure for analysis, creates a customizable run script for Popoolation2 for all pairwise comparisons, runs Popoolation2 and poolfstat, imports Popoolation2 and poolfstat output, and finally, generates population genetic statistics and plots for data visualization.

## Filter Raw Sequences
This script uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
For example, the default length of reads is 20 bp (so if you read is less than 20 bp then the read will be removed). Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp. 
The --max_n flag specificies how many "Ns" can be in the read before the read if discarded as "poor quality". Again, depending on your sequencing depth you may want to be more or less conservative. There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded). This is probably more conservative than you need to be, becasue the read may still contain infromative sequencign data.
Please note that these parameters are not specific to a Pool Seq dataset.
```{bash eval = FALSE}
#!/bin/bash

@@

## Pop Gen and Downstream Statistical Analyses
These in depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help

You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations. Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why its a good idea to have an estimate of your mean/min/max coverage for each sample/population. The minimum count must be less than the minimum coverage. You may want to modify this flag based on the number of inidividuals pooled per sample/population.

@@

***NOTE ABOUT MAX COVERAGE***
When genotyping individually barcoded samples, max coverage requirements are generally not recommended. However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc, because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates. PCR duplicates from a single individual in the pool may skew allele frequency estimates.

@@

--min-count: the minimum count of the minor allele, used for SNP identification. SNPs will be identified considering all samples/populations simultanously. default=2. The minimum count MUST be smaller than the min-coverage flag for this script to run.

@@

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the --window-summary-method flag can be used to determine pvalue for windows of SNPs. The default is to multiply the pvalue of all SNPs for the window. Alternatively, you can specify how pvalues within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependnt on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expect to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from idenitfying the true targets of selection, and therefore I use the geometric mean option.

@@

## Cochran-Mantel-Haenszel (CMH) test
This script will detect consistent allele frequency changes in biological replicates.
In my case, I was comparing selected samples to one control sample, but if you have replicates of treatments and controls, you can compare columns 1-2, 3-4, etc…

@@

## PoPoolation Pipelines for analyses
## Nucleotide Diversity
Calulate Pi (nucleotide diversity) for individual samples. Use mpileup files that were created for a single sample as the input. Below is an example for one sample.
```{bash eval=FALSE}
#!/bin/bash

>>>>

I mostly followed the PoPoolation and PoPoolation2 pipelines by Robert Kofler and Christian Schlötterer (https://sourceforge.net/p/popoolation2/wiki/Main/). The website has an easy-to-follow tutorial for using PoPoolation2 and how to use their scripts. This pipeline extends the use of the scripts for running on a Linux cluster. I highly recommend following the PoPoolation2 tutorial and manual and using these scripts if you have any modifications from default flags.
*Note: These scripts were written for use on Louisiana State University’s high performance cluster “SuperMike-II” running Red Hat Enterprise Linux 6 (more info found here: http://www.hpc.lsu.edu/resources/hpc/system.php?system=SuperMike-II)

The following scripts were used to analyze data from an experimental evolution study on *Tigriopus californicus* copepods (Griffiths JS, Kawji Y, Kelly MW. 2020. An experimental test of adaptive introgression in locally adapted populations of splash pool copepods. Mol Biol and Evol. 38(4):1306-1316. https://doi.org/10.1093/molbev/msaa289). Raw sequencing data are deposited in NCBI's Short Reads Archive (https://www.ncbi.nlm.nih.gov/bioproject/PRJNA597336/). To download this dataset, go to: https://trace.ncbi.nlm.nih.gov/Traces/sra/sra.cgi?view=run_browser , then type in the accession number for the identified files you want to download (e.g., the accession no. for the file Bodega population file is SRR10760095, identified under the "Run" heading). Then click on the Data access tab and there will be a downloadable link. Additional analyses and scripts can be found at: https://github.com/JoannaGriffiths/Tigriopus_HER

***Strengths and weaknesses of Pool-seq***
Due to the nature of Pool-seq where multiple individuals are pooled to prepare a single library, this results in a loss of haplotype information. Therefore, analyses rely on allele frequency estimates rather than hard “genotype calls”. PoPoolation2 is primarily used for comparing allele frequency differences between populations or treatments in comparison to a reference genome. The pooling of many individuals can be a cost-effective tool to character genetic variation present at the population level (e.g., genome wide association studies and experimental evolution). For example, to identify small changes in allele frequencies, you would need to genotype hundreds of individuals across the entire genome, which may be too cost-prohibitive with individual barcoding. However, individual genotype information is lost, as well as information on dominance and
effect sizes. For example, individual genotypes cannot be linked to individual phenotypes. Additionally, estimations of linkage disequilibrium are limited to a single sequencing read since that is the known limit of haplotype information for a single individual. Finally, distinguishing between sequencing errors and rare, low-frequency alleles is limited. Unlike sequencing of individuals, this cannot be solved by analyzing multiple reads from the same region of a single chromosome. One way to ameliorate this issue is to sequencing replicates of pools. In addition, Pool-seq analyses can be prone to inaccurate estimations of allele frequencies based on the experimental methods. Care must be taken to ensure the the experimental design maximizes the ratio sequencing coverage to the number of individuals pooled.

***Notes on Experimental Design for Downstream Analyses:***
It is important to try to standardize the number of individuals (and thus DNA) pooled for each sample. Robert Kofler and team also recommend that the pool size should be larger than the goal coverage for sequencing, because this minimizes re-sampling the same allele from a single individual several times. Because PoPoolation is designed for comparing differences in allele frequencies among treatments/population, accuracy will depend on the pool size and sequencing coverage. For example, smaller differences in allele frequency changes can be seen if you have a large number of individuals in a single pool and more coverage. The PoPoolation2 Manual page provides the following calculation to detect significant allele frequency differences: "To detect a significant (Fisher's exact test; p=0.01) allele frequency difference between two populations of 30%, a coverage (and pool size) of 50 will be sufficient. If however significant (Fisher's exact test; p=0.01) allele frequency differences of 10% need to be detected the coverage as well as the pool size should be about 400."
The sequencing accuracy of Pool-seq increases with larger pool sizes; if the number of sequenced individuals is approximately equal to the coverage, then the sequency accuracy is no longer superior to an individual level of sequencing (see Fig. 1a from Schlötterer et al., 2014).

@@

Customizable Filtering and Assessing Quality of Pool-Seq Data:
https://github.com/ToBoDev/assessPool
This pipeline filters SNPs based on adjustable criterion with suggestions for pooled data. It determines pool number and prepares proper data structure for analysis, creates a customizable run script for Popoolation2 for all pairwise comparisons, runs Popoolation2 and poolfstat, imports Popoolation2 and poolfstat output, and finally, generates population genetic statistics and plots for data visualization.

## Filter Raw Sequences
This script uses the wrapper TrimGalore to remove Adapter/Barcode sequences and removes poor quality reads. 
This example dataset uses paired end sequencing so I used the flag (--paired) in TrimGalore. Because these example files had high coverage, I was able to use conservative filters. 
For example, the default length of reads is 20 bp (so if you read is less than 20 bp then the read will be removed). Since I had high coverage and long read lengths (150 bp), I used a minimum read length of 40 bp. 
The --max_n flag specifies how many "Ns" can be in the read before the read if discarded as "poor quality". Again, depending on your sequencing depth you may want to be more or less conservative. There is no default value specified in TrimGalore, but I used a highly conservative value of 1 (ie any instances of N in the read and that read will be discarded). This is probably more conservative than you need to be, because the read may still contain informative sequencing data.
Please note that these parameters are not specific to a Pool Seq dataset.
```{bash eval = FALSE}
#!/bin/bash

@@

## Pop Gen and Downstream Statistical Analyses
These in-depth notes are not on the tutorial or manual page, but detailed descriptions for each flag can be found using the --help option when running each script. e.g. perl fst-sliding.pl --help

You will want to play around with the min-coverage, max-coverage, and min-count based on the mean coverage of your samples/populations. Minimum coverages that are too high and max coverages that are too low may result in an empty output file, which is why its a good idea to have an estimate of your mean/min/max coverage for each sample/population. The minimum count must be less than the minimum coverage. You may want to modify this flag based on the number of individuals pooled per sample/population.

@@

***NOTE ABOUT MAX COVERAGE***
When genotyping individually barcoded samples, max coverage requirements are generally not recommended. However, pool-seq analyses require a max coverage for calculating Fst, cmh, allele frequency changes, etc. because allele frequency estimates are based on the number allele count, which are prone to PCR duplicates. PCR duplicates from a single individual in the pool may skew allele frequency estimates.

@@

--min-count: the minimum count of the minor allele, used for SNP identification. SNPs will be identified considering all samples/populations simultaneously. default=2. The minimum count MUST be smaller than the min-coverage flag for this script to run.

@@

## Calculate Fisher's Exact Test
This script will calculate significant allele frequency differences for each SNP position between two samples. Results can be plotted in a Manhattan plot. 
Depending on the experimental method, the --window-summary-method flag can be used to determine pvalue for windows of SNPs. The default is to multiply the pvalue of all SNPs for the window. Alternatively, you can specify how pvalues within a window should be summarized. Options include multiple, geometric mean, or median.  The summary of p-values within a window is dependent on the experimental design and how recent selection was expected to occur. For example, the default parameter might be appropriate if your treatment comparisons are geographically separated populations, where selection is expected to have occurred over hundreds of thousands of years. However, this might be inappropriate for short-term experimental evolution studies, such as the example dataset here. In this dataset, copepods were selected in the lab for 15-20 generations, therefore, we expect relatively large linkage blocks that prevent us from identifying the true targets of selection, and therefore I use the geometric mean option.

@@

## Cochran-Mantel-Haenszel (CMH) test
This script will detect consistent allele frequency changes in biological replicates.
In my case, I was comparing selected samples to one control sample, but if you have replicates of treatments and controls, you can compare columns 1-2, 3-4, etc.

@@

## PoPoolation Pipelines for analyses
## Nucleotide Diversity
Calculate Pi (nucleotide diversity) for individual samples. Use mpileup files that were created for a single sample as the input. Below is an example for one sample.
```{bash eval=FALSE}
#!/bin/bash
